Iterative Monomorphisation

### 2024-07-08 iterative_monomorphisation.tex
related work

The alternative is to translate polymorphic problems to monomorphic problems. One approach is to encode polymorphism using dedicated function or predicate symbols \cite{mono-trans} in a monomorphic logic. Another approach is to encode polymorphism using {iterative monomorphisation}, as introduced by B\"ohme \cite[Section 2.2.1]{sb-phd}. This method relies on heuristically instantiating the formulae's type variables with concrete types.

@@

B\"ohme's iterative approach is implemented as part of the SMT (satisfiability modulo theories) integration \cite[Chapter 2]{sb-phd} in Isabelle/HOL. This implementation is also used by Sledgehammer \cite{judgement, hammer} to interface with superposition based automatic theorem provers. However, it is is documented only as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.

@@

% target page number: 0.5-1
\section{Related work}
\label{sec:related-work}

  * monomorphisation:
    Isabelle for \textit{smt} tactic (Böhme), \textit{metis}, Sledgehammer
    Meson of HOL Light

  * explicitly quantified formulas a la TPTP, existentials, how to eliminate

  * Böhme

>>>>

The alternative is to translate polymorphic problems to monomorphic problems.
One approach is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic;
we refer to Blanchette et al.\ \cite[Section~9]{mono-trans} for an overview of this approach.
Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme \cite[Section 2.2.1]{sb-phd}. This method relies on heuristically instantiating the formulae's type variables with concrete types.

@@

B\"ohme's iterative approach is implemented as part of the SMT (satisfiability modulo theories) integration \cite[Chapter 2]{sb-phd} in Isabelle/HOL. This implementation is also used by Sledgehammer \cite{judgement, hammer} to interface with superposition based automatic theorem provers. However, it is is documented only as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.
A similar algorithm appears to be implemented in the \texttt{MESON} tactic \cite{harrison-1996} of HOL Light, but it is undocumented.

@@

% target page number: 0.5-1
%\section{Related work}
%\label{sec:related-work}
%
%  * monomorphisation:
%    Isabelle for \textit{smt} tactic (Böhme), \textit{metis}, Sledgehammer
%  * a precursor for this work is the \verb|POLY_ASSUME_TAC| preprocessor of
%    MESON for HOL Light. It is unfortunately not documented beyond the comment
%    ``Push duplicated copies of poly theorems to match existing assumptions''
%    but it appears to perform at least partial monomorphisation.
%
%  * alternative: encode polymorphism
%    * many approaches to encode monomorphic and polymorphic types, starting
%      with Enderton \cite{xxx}, Stickel \cite{xxx}, and Wick and McCune
%      \cite{xxx}
%    * Blanchette et al.~\cite{xxx} exploit ``monotonicity'' to erase type
%      information
%    * see Blanchette et al.~\cite[Section~9]{xxx} for a more detailed coverage
%      of encodings of polymorphism
%    
%%\li{https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette/enc_types_article.pdf}
